---
title: "Probability for Statistics / Elements of Probability --- Lab 7 Solutions"
output:
  pdf_document: default
  html_document:
    theme: yeti
---

In this lab, we will repeat Questions 6 and 8 from Tutorial 6 and some questions from Tutorial 7.


## Question 1 (T6/Q6)

If $X$ has a gamma distribution with $\alpha = 2$ and $\theta = 4$, find $P(X < 5)$.

```{r}
alpha = 2; theta = 4
pgamma(5, alpha, scale = theta)
```


***

## Question 2 (T6/Q8)

Let $X$ have a $\chi^2(2)$ distribution. Find constants $a$ and $b$ such that:

$P(a<X<b)=0.90$ and $P(X<a)=0.05$

```{r}
df = 2

# Note: The second statements means that a must be the 5th percentile, and so, with the 
# first statement, we can conclude that b is the 95th percentile.

# a
qchisq(.05, df)

#b
qchisq(.95, df)
```


***

## Question 3 (T7/Q1)

For this question, we require the MosaicCalc package.

```{r, message=FALSE, warning=FALSE}
# This code installs the package only if it is not already installed, and loads it.
if(!("mosaicCalc" %in% installed.packages()[,"Package"])){
  install.packages("mosaicCalc", repos = "https://cran.ms.unimelb.edu.au/")
} 
library("mosaicCalc")
```

Let $X$ have the pdf $f(x) = 4x^3, 0<x<1$. Find and plot the pdf of $Y=X^2$ using the change of variable technique. You should define four functions throughout the process: $f(x)$, $v(y)$, $v'(y)$, and the pdf of $Y$.

```{r}
# The support of Y is (0,1)
# X = sqrt(Y) = v(Y)

# Y = u(X) = X^2 is a continuous increasing function on (0,1), so we use the change of 
# variable technique:

v <- function(y) y^0.5
f <- function(x) 4*x^3
dv <- D(v(y)~y)
f.y <- function(y) f(v(y))*abs(dv(y))
curve(f.y, from=0, to=1, xlab = "y", ylab = "P(Y=y)", main="pdf of Y")
```

This plot confirms that the pdf of $Y$ is $f_Y(y) = 2y$.


***

## Question 4

*This question relates to Question 4 of Tutorial 7 (T7/Q4). Ensure that you have completed it before attempting this.*

### (a)

In Part (a) of T7/Q4 you found the marginal pmf of $X$ to be

$f_X(x)=\frac{4x+10}{32},\quad x=1,2.$

Define this as a function, then use the sample(...) function to generate 10000 observations from this marginal pmf. Assign the results to a variable.

```{r}
pmf.x <- function(x) (4*x+10)/32
x_vals <- c(1,2)
n <- 10000

x_results <- sample(x = x_vals, n, replace = T, prob = pmf.x(x_vals))
```


### (b)

In Part (b) of T7/Q4  you found the marginal pmf of $Y$ to be

$f_Y(y)=\frac{3+2y}{32},\quad y=1,2,3,4.$

Define this as a function, then use the sample(...) function to generate 10000 observations from this marginal pmf. Assign the results to a variable.

```{r}
pmf.y <- function(y) (3+2*y)/32
y_vals <- c(1,2,3,4)
n <- 10000

y_results <- sample(x = y_vals, n, replace = T, prob = pmf.y(y_vals))
```


### (c)

Calculate the proportion of times where, for a given observation $i$, $X>Y$.

```{r}
sum(x_results > y_results)/n
```


### (d)

Calculate the proportion of times where, for a given observation $i$, $Y=2X$. 

```{r}
sum(y_results == 2*x_results)/n
```


### (e)

Calculate an estimate of $E(X)$ using the sample mean. 

```{r}
mean(x_results)
```


### (f)

Calculate an estimate of $E(X+Y)$ using the results of the $X$ and $Y$ simulations. 

```{r}
mean(x_results+y_results)
```


### (g)

The large number $n=10000$ of observations should lead to estimates that are quite accurate. Which of the above calculated values match up with their equivalent answers for Parts (c), (d), (h) and (i) of T7/Q4? Why/Why not?

**My answer:**

The answers for our estimates of $E(X)$ and $E(X+Y)$ in Parts (e) and (f) above (the estimated means) respectively match with Parts (h) and (i) of T7/Q4. Note that $X$ and $Y$ are not independent, as per our answer to Part (g) of T7/Q4. Parts (h) of of T7/Q4 makes no assumption on $X$'s independence from $Y$, while for Parts (i) of of T7/Q4, $E(X+Y) = E(X) + E(Y)$ irrespective of whether $X$ and $Y$ are independent; this justifies the accuracy of the estimates.

However, our answers in Part (c) and (d) above do not match with Parts (c) and (d) of T7/Q4. This is because simulating $X$ and $Y$ *independently* using their marginal pmfs and then comparing them directly assumes that $X$ and $Y$ are independent.


***

### Question 5

In Module 4, we learned about the trinomial distribution, an extension of the binomial distribution where 3 outcomes are possible for each trial. This, in turn, can be generalised to a *multinomial distribution*, where $k\geq2$ outcomes are possible for each trial. The binomial distribution corresponds to the case where $k=2$, and the trinomial distribution corresponds to the case where $k=3$.

R has the functionality to generate vectors of multinomially distributed random values and compute multinomial probabilities. Simply search up "Multinomial" in the Help window to see how to use the available functions, or type ?Multinomial in the Console.

### (a)

Let $(X,Y)\stackrel{d}{=} Trinom(n=19, p_x=0.4, p_y=0.2)$. 

To find $P(X=3, Y=6)$ we type the command:

```{r}
dmultinom(c(3,6,10), 19, c(0.4,0.2,0.4))
```

Find $P(X=3, Y\leq 6)$. 

*You may need to use a 'for loop' to sum the probabilities of $Y$ taking values less than or equal to six.*

```{r}
sum <- 0
for (i in 0:6) {
  # same as above, the second outcome value is i, the third is 19-3-i = 16-i
  sum = sum + dmultinom(c(3,i,16-i), 19, c(0.4,0.2,0.4))
}
sum
```


### (b)

Perform a simulation of $(X,Y)$. Note the results of the simulation, that is, the outcome of $X$ and $Y$.

```{r}
# Note the first parameter n equals 1 because we are performing 1 simulation
rmultinom(1, 19, c(0.4,0.2,0.4))
```

**My answer:**

The first value is the outcome of $X$, the second is the outcome of $Y$.


### (c)

Let $B \stackrel{d}{=} b(6, 0.35)$ (binomial). Find $P(B=2)$ first using the dbinom(...) function, then using the dmultinom(...) function.

```{r}
dbinom(2,6,0.35)
# OR
dmultinom(c(2,4), 6, c(0.35,0.65))
```


### (d)

Hogwarts is hosting the Triwizard Tournament this year. Lily Luna Potter of Gryffindor is representing Hogwarts, Dominique Weasley is representing the Beauxbatons Academy of Magic, and Luka Krum is representing the Durmstrang Institute. According to the Wizardsbet, Potter has a 40% chance of winning each of the three unknown events, Weasley a 15% chance, and Krum 45%. What is the probability that all competitors will win one event each?

```{r}
dmultinom(c(1,1,1), 3, c(0.4, 0.15, 0.45))
```


### (e)

Mia is at a family Christmas lunch, where there are 20 bonbons around the table. The bonbons hold different sorts of prizes --- in general, 50% of bonbons hold paper crowns, 30% hold keyrings, and 20% hold chocolates. Mia, who doesn't like the smell of the lunch her family is cooking up, would prefer to fill up on the bonbon chocolates instead.

She decides that she will challenge every attendee to their bonbon. Mia has a 56% chance of winning a bonbon duel --- she researched a way to cheat prior to the lunch.

Simulate the number of chocolates Mia will win.

```{r}
num_won <- rbinom(1, 20, 0.56)
rbinom(1, num_won, 0.2)
```
