---
title: "Probability for Statistics / Elements of Probability --- Lab 6 Solutions"
output:
  pdf_document: default
  html_document:
    theme: yeti
---


## Question 1

Suppose a value $x$ is chosen “at random” in the interval [0,8]. In other words, $x$ is an observed value of a random variable $X \stackrel{d}{=} U(0,8)$. The random variable $X$ divides the interval [0,8] into two subintervals, the lengths of which are $X$ and $8-X$, respectively. Denote by $Y = \min(X,8-X)$, the length of the shorter one of the two intervals. 

### (a)

Conduct a simulation to generate 100000 observations of $Y$.

*Hint: Enter ?runif and ?pmin in the Console to see how these functions may be useful.*

Plot the results in a histogram.

```{r}
x <- runif(100000,0,8)
eight_min_x <- 8-x
y <- pmin(x, eight_min_x)

hist(y, freq=FALSE, main="Results of Simulation of Y")
```


### (b)

What is the cdf of $Y$, and the name of its distribution?

**My answer:**

From the simulations and the shape of the resulting histogram, the random variable $Y$  would be uniformly distributed $Y \stackrel{d}{=} U(0,4)$. This can indeed be checked analytically (exercise). Therefore the cdf of $Y$ is $F(y) = 0.25y, 0 \leq y \leq 4$.


***

## Question 2

For this question, we require the MosaicCalc package.

```{r, message=FALSE, warning=FALSE}
# This code installs the package only if it is not already installed, and loads it.
if(!("mosaicCalc" %in% installed.packages()[,"Package"])){
  install.packages("mosaicCalc", repos = "https://cran.ms.unimelb.edu.au/")
} 
library("mosaicCalc")
```

This question is based on Question 4 from Tutorial 6.

Let the random variable $X$ have the pdf

$$f(x)=\frac{e^{-x}}{(1+e^{-x})^2},\quad -\infty<x<\infty.$$




### (a)

Plot the pdf of $X$ from $x=-10$ to $x=10$. What does the shape of this pdf indicate about the mean of $X$?


```{r}
pdf <- function(x) exp(-x)*(1+exp(-x))^-2
curve(pdf(x), from=-10, to=10, ylab="P(X=x)", main="pdf of X")
```

**My answer:**

It is an even function, centered around $x=0$. Therefore, the mean will be 0.


### (b)

Plot the cdf of $X$ from $x=-10$ to $x=10$.

*Hint: Use the imported antiD(...) function from the package MosaicCalc. If the error "non-finite function value" appears when approximating -Inf by a large negative number, try using -10^2.*

```{r}
antiD_pdf <- antiD(pdf(x)~x)

# Remember to subtract the lower terminal
cdf <- function(x) antiD_pdf(x)-antiD_pdf(-10^2)

curve(cdf(x), from=-10, to=10, ylab="P(Y<=y)", main="cdf of X")
```


### (c)

Find the mean and variance of $X$. Does the mean correspond to your answer for Part (a)?

*Hint: If you would like the integrate(...) function to return only the numerical approximation of the integral (without the absolute error), you need to use integrate(...)$value.*

*Note: For this question, if the error "non-finite function value" appears when using the integrate(...) function with terminals -Inf and Inf, try using -+10^6, or -+10^2.*

```{r}
integrand1 <- function(x) pdf(x)*x
mean <- integrate(integrand1,-Inf,Inf)$value

integrand2 <- function(x) pdf(x)*x^2
Ex2 <- integrate(integrand2, -10^2, 10^2)$value

var <- Ex2 - mean^2

mean
var

# mean = 0, var = pi^2/3
```

**My answer:**

Yes, the mean is zero, with a negligible error.


### (d)

Find $P(3<X<5)$ using the pdf.

```{r}
integrate(pdf,3,5) 
```


### (e)

Find $P(3<X<5)$ using the cdf.

```{r}
cdf(5)-cdf(3)
```


### (f)

Let $Y=\frac{1}{1+e^{-X}}$. Plot the cdf of $Y$ using the distribution function technique covered in the lectures.

```{r}
# Note the space of Y is 0 < y < 1
# We have P(Y<=y) = P(1/(1+exp(-X)) <= y) = P(X <= log(y/(1-y)))

y.cdf <- function(y) cdf(log(y/(1-y)))
curve(y.cdf, from = 0.01, to = 0.99, xlab="y", ylab="P(Y<=y)", main="cdf of Y")
```

Can you tell the name of the distribution of $Y$?

**My answer:**

$Y$ has a Unif(0,1) distribution.


*** 

## Question 3

Let $X$ have an Exponential distribution with mean $w>0$. Assume $u$ and $v$ are two positive real numbers.

### (a)

Create a function which, for given values of $u$ and $w$, finds $P(X>u)$.

```{r}
# Note that rate = 1/w
u.fun <- function(u,w) 1 - pexp(u, 1/w)
```


### (b)

Create a function which, for given values of $u$, $v$ and $w$, finds $P(X>u+v)$.

```{r}
uv.fun <- function(u,v,w) 1 - pexp(u+v, 1/w)
```


### (c)

Create a function which, for given values of $u$, $v$ and $w$, finds $P(X>u+v|X>v)$.

```{r}
given.fun <- function(u,v,w) uv.fun(u,v,w)/u.fun(v,w)
```


### (d)

Compare the outputs of the functions from Parts (a) and (c) when given specific values of $u$, $v$ and $w$.

```{r}
u = 1; v = 1; w = 1
u.fun(u,w)
given.fun(u,v,w)

u = 2.5; v = 0.8; w = 3
u.fun(u,w)
given.fun(u,v,w)

u = 12; v = 0.1; w = 18
u.fun(u,w)
given.fun(u,v,w)
```


What conclusion can you draw? What is the reason behind the conclusion?

**My answer:**

No matter the values of $u>0$, $v>0$ or $w>0$, the results of $P(X>u)$ and $P(X>u+v | X>v)$ are equal.

This is because the exponential distribution has the *no-memory property*.


***

## Question 4

This question illustrates the relationship between Gamma, Chisquare and Poisson distributions.

Let $X$ be a Gamma random variable with shape parameter $\alpha=8$ and scale parameter $\theta=2$. That is, the pdf of $X$ is:

$$f(x)=\frac{1}{\Gamma(8)2^8}x^7e^{-x/2}=\frac{1}{1290240}x^7e^{-x/2},\quad x\geq 0.$$
*Hint: Type ?pgamma, ?ppois and ?pchisq in the Console to see how to calculate probabilities involving Gamma, Chisquare and Poisson distributions.*

### (a)

Compute $P(X>3)$.

```{r}
alpha = 8; theta = 2
1 - pgamma(3, alpha, scale = theta)

```


### (b)

Create a Poisson random variable $Y$ with parameter $\lambda=\frac{3}{2}$.

Then compute the probability $P(Y\leq 7)$.


```{r}
ppois(7, lambda = 3/2)
```

Compare the result with $P(X>3)$ in Part (a). Why are they the same?

**My answer:**

We see that $P(X>3)=P(Y\leq 7)$. The reason is that the event that the `waiting time' for the 8th change is greater than 3 is the same as the event that there are at most 7 changes during 3 time units.


### (c)

Create a chi-square random variable $Z$ with degrees of freedom 16. Compute $P(Z>3)$.

```{r}
df = 16
1-pchisq(3,df)
```

Is this equal to $P(X>3)$? Why?

**My answer:**

We see that $P(X>3)=P(Z>3)$. The reason is that the chi-square distribution with $r$ degrees of freedom corresponds to a Gamma distribution with $\theta=2$ and $\alpha=r/2$.

